---
title: 'Lab 5: HPV Vaccination Completion Rates (Logistic Regression)'
author: "Merhan Ghandehari"
date: "March 13, 2016"
output: word_document
---
# Introduction
Gardasil is a vaccine that has recommended for all women aged 9–26 in order to be protected from HPV (human papillomavirus) virus. This vaccin should be given three times within a 6–12 month timespan. In this research we are going to investigate the effects of different factors that can result in failure to complete the three-shot sequence. That is, we want to find “good” predictors for regimen completion in order to answer the following question:

- Which groups of patients appear to have a higher rate of completion? (this groups are defined by age, race and urban or suburban clinics)
- Which variables and patient characteristics (e.g., location of clinic, insurance type, and type of practice) best predict Gardasil vaccination completion?

Theoretically race and socioeconomic status can have a strong effect on the completion rate. For example, it seems that women who leave in poor, minority-heavy communities and among those lacking health insurance are less likely to complete the three-shot sequence within the 12-month period. So patient demographics, socioeconomic status, and care physician characteristics are some factors that can be examined in this research. Also we want to examin this hypotheis that whether patients who receive medical assistance and/or go to urban clinics are more likely to fail to complete the regimen than those who have some sort of insurance and/or go to suburban clinics.

# Methods

Our dependent variable in this lab (i.e., completion of the HPV vaccination) was a binary variable (yes/no outcome). So, regression model was used as an appropriate method for modeling a binary dependent variable. All of the independent variales, except age, were categorical and we converted them to factor. In addition, we created dummy variables bacuase our categorical variabels have several sub-categories, and we are interested to to assess each subcategory and if significant use it in the multivariate logistic regression. Also we noticed that there is a descrepency between the variable "completed"" and "number of shots". So we decided to correct the completed vaiable based upon the number of shots that the patients completed.

First of all the impact of our categorical variables, as independent variables, on vaccination completion  was examined by using tables, plots, and running a simple logistic regression. For each categorical variable, we created a table (e.g., `table(gardasil$Completed,gardasil$Location)`), and then we calculated the percentages for each column (`prop.table(locs, 2) *100`). This table helps a lot to find the most significant elements of each variable. After making simple logistic regression, probabilities, odds ratios and their confidence interval were calculated and assessed. Also anova test was run. I also ploted the fitted values (e.g., `plot(gardasil$Location, fitted(glmLocation)`) that was quite useful to find what element of each variable is statisticly significant. I did not run a univariate logistic regression on each individual dummy variable, because it was quite streightforward to find the most effective elements of each group based on the above-mentioned methods and I would explian the details in the result section.  

After finding the significant variables, I fit a multivariate logistic regressions to the most interesting variables I found in the previous step. First I compared this model to the simple regression models. Although some of the variables seems significant in the context of a simple model, but they are not signficant in a full model. So we can conclude that a simple model cannot explain the whole variation in out dependent variable. Then I eliminated the non-significant variables and assesed the reduced moldels. I used `AIC`, `anova(fit.reduced1, test = "LRT")`, `drop1(fit.reduced2, test = "LRT")`, and `anova(fit.reduced1, fit.reduced2, test = "Chisq")` to assess the models. I make four reduced model to arrive at my final model.

Finally I examined my model for any possible interaction affects. I tested individual interactions to see
which one improve the model using summray of logistic model (AIC) and annova test. Then I added some interaction terms to my model. 

```{r echo=FALSE, results='hide', message=FALSE, eval=F}
# here is examples to show the main parts of my model workflow
##############Make Dummy Variables###################
#example
# Dummy Variables: Age Group
gardasil$Yrs11_17 = ifelse(gardasil$AgeGroup == 0, c(1), c(0)) #11-17 years = 1 in Dummy
gardasil$Yrs18_26 = ifelse(gardasil$AgeGroup == 1, c(1), c(0)) #18-26 years = 1 in Dummy

###Converting the categorical variables to factor########

for (g in which(names(gardasil)!= "Age")){
  gardasil[,g] = as.factor(gardasil[,g])
}


#########assign names to the dataframe###############
gardasil$MedAssist = factor(gardasil$MedAssist, 
                             levels=c(0,1),  
                             labels=c("NO_Asst", "Asst")) 

################ Bivariate Logistic Regression Models#################
# example
#completion rates by location 
locs= table(gardasil$Completed,gardasil$Location) 
kable(prop.table(locs, 2) *100, digits = 1) # column percentages
barplot(prop.table(locs, 2), beside=TRUE)
glmLocation=glm(gardasil$Completed~gardasil$Location, family=binomial) 
kable(exp(cbind(OR=coef(glmLocation),confint(glmLocation))), digits = 4)
summary(glmLocation)
anova(glmLocation, test = "LRT")
plot(gardasil$Location, fitted(glmLocation), 
     main = "Probability of Vaccine Regimen Completion By Location", 
     xlab = "Location", ylab = "Gardasil Completion Probability")

##############Multivariate Logistic Regression Models###################
fit.full = glm(Completed ~ MedAssist + Black + LocationType + Pediatric + PrivatePayer + gardasil$Age + 
                HospitalBased + Hispanic + FamilyPractice, data=gardasil, family="binomial")
summary (fit.full)
anova(fit.full, test = "LRT")
lmtest:::lrtest(fit.full)

fit.reduced1 = glm(Completed ~ MedAssist + Black + LocationType + gardasil$Age + 
                HospitalBased + FamilyPractice + Pediatric, data=gardasil, family="binomial")
summary (fit.reduced1)
anova(fit.reduced1, test = "LRT")
anova(fit.full, fit.reduced1, test = "Chisq")
drop1(fit.reduced1, test = "LRT")
```

# Results:

```{r echo=FALSE, results='hide', message=FALSE}
# Load Packages and Libraries:
library(boot)
library(car)
library(MASS)
library(knitr)
library(leaps)
```

```{r echo=FALSE, results='hide',message=FALSE}
# Load Data
gardasil = read.table("~/Desktop/Quantitative/Lab5/jh_gardasil.dat", header=TRUE)

# we noticed that there are a descrepency between the variable completed and number of shots. So we decided # to correct the completed vaiable based upon the number of shots that the patients completed.
gardasil$Completed = ifelse(gardasil$Shots == 3 ,1, gardasil$Completed)

# Summarize Data 
summary(gardasil)

```

```{r echo=FALSE, results='hide',message=FALSE}
##############Make Dummy Variables###################
# in order to use subcategories in each vaibale as independent vaiables in our regression we created dummy vaiable
# for each sub-category.

# Dummy Variables: Age Group
gardasil$Yrs11_17 = ifelse(gardasil$AgeGroup == 0, c(1), c(0)) 
gardasil$Yrs18_26 = ifelse(gardasil$AgeGroup == 1, c(1), c(0)) 

# Dummy Variables: Race
gardasil$White = ifelse(gardasil$Race == 0, c(1), c(0))
gardasil$Black = ifelse(gardasil$Race == 1, c(1), c(0))
gardasil$Hispanic = ifelse(gardasil$Race == 2, c(1), c(0))
gardasil$Other = ifelse(gardasil$Race == 3, c(1), c(0))

# Dummy Variables: Insurance Type
gardasil$MedicalAssist = ifelse(gardasil$InsuranceType == 0, c(1), c(0))
gardasil$PrivatePayer = ifelse(gardasil$InsuranceType == 1, c(1), c(0))
gardasil$HospitalBased = ifelse(gardasil$InsuranceType == 2, c(1), c(0))
gardasil$Military = ifelse(gardasil$InsuranceType == 3, c(1), c(0))

# Dummy Variables: Location
gardasil$Odenton = ifelse(gardasil$Location == 1, c(1), c(0))
gardasil$WhiteMarsh = ifelse(gardasil$Location == 2, c(1), c(0))
gardasil$JohnsHopkins = ifelse(gardasil$Location == 3, c(1), c(0))
gardasil$Bayview = ifelse(gardasil$Location == 4, c(1), c(0))

# Dummy Variables: Practice Type
gardasil$Pediatric = ifelse(gardasil$PracticeType == 0, c(1), c(0))
gardasil$FamilyPractice = ifelse(gardasil$PracticeType == 1, c(1), c(0))
gardasil$OBGYN = ifelse(gardasil$PracticeType == 2, c(1), c(0))

# Check MedAssist matches MedicalAssist Insurance Type
ComparisonMedAsst = ifelse(gardasil$MedicalAssist == gardasil$MedAssist, c(0), c(1)) # Counts mismatches 
sum(ComparisonMedAsst) # If mismatch occurs, sum will be > 0. If not, the two are equal)
# Test determines MedAssist and MedicalAssist Insurance Type are the same.

```

```{r echo=FALSE, results='hide',message=FALSE}
# Converting the categorical variables to factor

for (g in which(names(gardasil)!= "Age")){
  gardasil[,g] = as.factor(gardasil[,g])
}


#assign names to the dataframe
gardasil$MedAssist = factor(gardasil$MedAssist, 
                             levels=c(0,1),  
                             labels=c("NO_Asst", "Asst")) 
gardasil$Location = factor(gardasil$Location, 
                            levels = c(1,2,3,4),  
                            labels=c("Odenton", "WhiteMarsh", "JohnsHopkins",  "Bayview")) 
gardasil$LocationType = factor(gardasil$LocationType, 
                                levels=c(0,1), labels=c("Suburban", "Urban")) 
gardasil$Completed = factor(gardasil$Completed, 
                             levels=c(0,1), labels=c("No", "Yes")) 
gardasil$Race = factor(gardasil$Race, 
                        levels=c(0,1,2,3),  
                        labels=c("White", "Black", "Hispanic", "Other"))
gardasil$InsuranceType = factor(gardasil$InsuranceType, 
                        levels=c(0,1,2,3),  
                        labels=c("MedAssis", "Private", "Hospital", "Military"))
gardasil$PracticeType = factor(gardasil$PracticeType, 
                                 levels=c(1,2,3),  
                                 labels=c("Pediatric", "FamilyPrac", "OB_GYN"))
gardasil$AgeGroup = factor(gardasil$AgeGroup, 
                                levels=c(0,1),  
                                labels=c("Yrs11_17", "Yrs18_26"))

# Summarize Data 
# here is a smmary of our data after creating dummy variables, Converting the categorical variables to factor and assigning names to the variables
summary(gardasil)

```

This section includes some of the R code, tables, and plots that best summarize the outputs.

1 - We calulated the total percentage of completed versus incompleted vaccines. The resulsts shows that `66%` of wemen have not gotten the full vaccination and only `33%` have succeded to complete the vaccination.

```{r results='hide'}

# percentage of completed vs incompleted vaccines
summary(gardasil$Completed) / sum(summary(gardasil$Completed))
#  No       Yes 
# 0.6680821 0.3319179 
```

2 - Here we have the Bivariate Logistic Regression Models for each variable:

2-1- completion rates by location 
```{r message=FALSE, fig.align='center'}

#completion rates by location 
locs= table(gardasil$Completed,gardasil$Location) 
kable(prop.table(locs, 2) *100, digits = 1) # column percentages
barplot(prop.table(locs, 2), beside=TRUE)
glmLocation=glm(gardasil$Completed~gardasil$Location, family=binomial) 
kable(exp(cbind(OR=coef(glmLocation),confint(glmLocation))), digits = 4)
#summary(glmLocation)
#anova(glmLocation, test = "LRT")
plot(gardasil$Location, fitted(glmLocation), 
     main = "Probability of Vaccine Regimen Completion By Location", 
     xlab = "Location", ylab = "Gardasil Completion Probability")

# Based on the results we can conclude that patients who go to the Johns Hopkins clinic have the lowest rate of completion and approximately half of the patients (highest completion rate) who go to the White Marsh complete their vaccination. Also we can see that the completion rates of clicnics where are located in urban are is much higher than the suburban clicnics.
```


2-2- Completion rates by Race  
```{r message=FALSE, fig.align='center'}

race= table(gardasil$Completed,gardasil$Race) 
kable(prop.table(race, 2) *100, digits = 1) # column percentages
barplot(prop.table(race, 2), beside=TRUE)
glmRace=glm(gardasil$Completed~gardasil$Race, family=binomial) 
kable(exp(cbind(OR=coef(glmRace),confint(glmRace))), digits = 4)
#summary(glmRace)
plot(gardasil$Race, fitted(glmRace), 
     main = "Probability of Vaccine Regimen Completion By Race", 
     xlab = "Race", ylab = "Gardasil Completion Probability")
# Based on the results, the race black is a significant predictor. The odds ratio of black is .54 and the probability of completion is the the lowest (~ 29%). 
```

2-3- Completion rates by Medical Assistance 
```{r message=FALSE, fig.align='center'}
MedAssist= table(gardasil$Completed,gardasil$MedAssist) 
kable(prop.table(MedAssist, 2) *100, digits = 1) # column percentages
barplot(prop.table(MedAssist, 2), beside=TRUE)
glmMedAssist=glm(gardasil$Completed~gardasil$MedAssist , family=binomial) 
kable(exp(cbind(OR=coef(glmMedAssist),confint(glmMedAssist))), digits = 4)
#summary(glmMedAssist)
plot(gardasil$MedAssist, fitted(glmMedAssist), 
     main = "Probability of Vaccine Regimen Completion by Medical Assistance", 
     xlab = "Medical Assistance", ylab = "Gardasil Completion Probability")

# It seems that the patients who use the medical assistace are less likely to finish the Gardasil regimen (probability of 25%). By the way, the probabilty that a person without Medical Assistance complete the vaccinatin is twice as a person with Medical Assistance

```

2-4- Completion rates by Location Type
```{r message=FALSE, fig.align='center'}
LocationType= table(gardasil$Completed,gardasil$LocationType) 
kable(prop.table(LocationType, 2) *100, digits = 1) # column percentages
glmLocationType=glm(gardasil$Completed~gardasil$LocationType , family=binomial) 
kable(exp(cbind(OR=coef(glmLocationType),confint(glmLocationType))), digits = 4)
#summary(glmLocationType)
plot(gardasil$LocationType, fitted(glmLocationType), 
     main = "Probability of Vaccine Regimen Completion By Location Type", 
     xlab = "Location Type", ylab = "Gardasil Completion Probability")

# Location type is also a significant predictor. Those who go to to the urban clinics are less likely finish the Gardasil regimen (probability of 29%). That is, the probablility of completion rate in a suburban clinic is double of a urban clinic.

```


2-5- Completion rates by Insurance Type
```{r message=FALSE, fig.align='center'}
InsuranceType= table(gardasil$Completed,gardasil$InsuranceType) 
kable(prop.table(InsuranceType, 2) *100, digits = 1) # column percentages
glmInsuranceType=glm(gardasil$Completed~gardasil$InsuranceType , family=binomial) 
kable(exp(cbind(OR=coef(glmInsuranceType),confint(glmInsuranceType))), digits = 4)
#summary(glmInsuranceType)
plot(gardasil$InsuranceType, fitted(glmInsuranceType), 
     main = "Probability of Vaccine Regimen Completion By Insurance Type", 
     xlab = "Insurance Type", ylab = "Gardasil Completion Probability")

# As we saw before Medical Assistance is a significant predictor. It seems that PrivatePayer is also can be considereded as a signficant predictor. The probability that PrivatePayer patients will complete the Gardasil regimen is about 38%. patients with hospital based insurance are significantly more likely to complete the Gardasil regimen than other insurance types.
```


2-6- Completion rates by Practice Type
```{r message=FALSE, fig.align='center'}
#completion rates by Practice Type
PracticeType= table(gardasil$Completed,gardasil$PracticeType) 
kable(prop.table(PracticeType, 2) *100, digits = 1) # column percentages
glmPracticeType=glm(gardasil$Completed~gardasil$PracticeType , family=binomial) 
kable(exp(cbind(OR=coef(glmPracticeType),confint(glmPracticeType))), digits = 4)
#summary(glmPracticeType)

# only 32% of the Pediatric group complete the vaccination. So, Pediatric group can be an intersting dummy variable to be incorporated in the multiple logistic regression. 
```


2-7- Completion rates by AgeGroup
```{r message=FALSE, fig.align='center'}
AgeGroup= table(gardasil$Completed,gardasil$AgeGroup) 
kable(prop.table(AgeGroup, 2) *100, digits = 1) # column percentages
glmAG=glm(gardasil$Completed~gardasil$AgeGroup, family=binomial) 
kable(exp(cbind(OR=coef(glmAG),confint(glmAG))), digits = 4)
#summary(glmAG)
plot(gardasil$AgeGroup, fitted(glmAG), 
     main = "Probability of Vaccine Regimen Completion By Age Group", 
     xlab = "Age Group", ylab = "Gardasil Completion Probability")

# The Probability of Gardasil Vaccine Regimen Completion does not appear to be significantly affected by AgeGroup. The Probability Decreases slightly with the older AgeGroup Aged 18-26. 
```


2-8- Completion rates by Age
```{r message=FALSE,message=FALSE, fig.align='center'}
Age= table(gardasil$Completed,gardasil$Age) 
kable(prop.table(Age, 2) *100, digits = 1) # column percentages
glmAge=glm(gardasil$Completed~gardasil$Age, family=binomial) 
kable(exp(cbind(OR=coef(glmAge),confint(glmAge))), digits = 4)
#summary(glmAge)
plot(gardasil$Age, fitted(glmAge), main = "Probability of Vaccine Regimen Completion By Age", 
     xlab = "Patient Age (in Years)", ylab = "Gardasil Completion Probability")
# Probability of completion does  appear to be affected by age. There is a negative (or inverse) relationship between completion of the Gardasil Vaccine Regimen and Age and it decreases slightly with age in this data. 

```

3- Multivariate Logistic Regression Models: We begin with a model with all of the parametes that were significant in our bivariate logistic regression models. 

```{r message=FALSE, fig.align='center'}
# Null hypothesis: the probability of a patient's vaccine completion is no better than the average probabiliy.
fit.full = glm(Completed ~ MedAssist + Black + LocationType + Pediatric + PrivatePayer + gardasil$Age + 
                HospitalBased + Hispanic + FamilyPractice, data=gardasil, family="binomial")
#summary (fit.full)
#anova(fit.full, test = "LRT")

#AIC: 1806.9

# because both Johns Hopkins and Bayview were significant and urban, I just used LocationType (urban = 1).

#From the p-values for the regression coefficients (last column), you can see that PrivatePayer and Hispanic may not make a significant contribution to the equation. This two variables are the least significant for both the Wald and LRT tests. We Will eliminate them in our next model.

#lmtest:::lrtest(fit.full)
#The chi-square of 89.825 with 9 degrees of freedom and an associated p-value of significantly less than 0.001 tells us that our model as a whole fits significantly better than an empty model. 


fit.reduced1 = glm(Completed ~ MedAssist + Black + LocationType + gardasil$Age + 
                HospitalBased + FamilyPractice + Pediatric, data=gardasil, family="binomial")
#summary (fit.reduced1)
#anova(fit.reduced1, test = "LRT")
#anova(fit.full, fit.reduced1, test = "Chisq")
# AIC: 1803
# Pediatric is the least significant for both the Wald and LRT tests. We Will eliminate them.

#The nonsignificant chi-square value (p = 0.9384) suggests that the reduced model fits as well as the full model, reinforcing our belief that PrivatePayer and Hispanic don’t add significantly to the prediction above and beyond the other variables in the equation. Therefore, you can base your interpretations on the simpler model.


fit.reduced2 = glm(Completed ~ MedAssist + Black + LocationType + gardasil$Age + 
                HospitalBased + FamilyPractice, data=gardasil, family="binomial")
#summary (fit.reduced2)
#anova(fit.reduced2, test = "LRT")
# AIC: 1803.6
# Each regression coefficient in this reduced model is statistically significant 
# The fit.reduced2 model has a similar AIC to mylogit2, so keep it.

#drop1(fit.reduced2, test = "LRT")
# Null hypothesis: There is no difference in the deviance between the two models.
#anova(fit.reduced1, fit.reduced2, test = "Chisq")
# All of the variables are sygnificant. So, we do not drop any variabels from our model. Also anova test using Chisq method illustrate that there not a sygnificant difference between model fit.reduced1 and fit.reduced2, So we keep the reduced model.

#interaction model
fit.reduced3 = glm(Completed ~ MedAssist + Black + Age + 
                 FamilyPractice + HospitalBased * LocationType, data=gardasil, family="binomial")
#summary (fit.reduced3)
#anova(fit.reduced3, test = "LRT")
#AIC: 1802.8
#anova(fit.reduced2, fit.reduced3, test = "Chisq")

fit.reduced4 = glm(Completed ~ MedAssist + Black + Age * LocationType + 
                     FamilyPractice + HospitalBased * LocationType, data=gardasil, family="binomial")
summary (fit.reduced4)
anova(fit.reduced3, test = "LRT")
kable(exp(cbind(OR=coef(fit.reduced4),confint(fit.reduced4))), digits = 4)
# AIC: 1781.3
# The fit.reduced4 model has the smallest AIC. Adding the (HospitalBased * LocationType) interaction term significantly improve the model. We can also conclude that there is relationship between Age and location, as well as between hospital based inssurance and location.
# This model shows that Race and Insurance Type are good predictors of Gardasil completion
# Based on our final model, we can say that MedAssist, Black, Age, LocationType, FamilyPractice, and HospitalBased are the variables and patient characteristics that can best predict Gardasil vaccination completion.
```
  
# Discussion:

The results of a bunch of univariate and multivariate logistic regressions illustarates that the probability (likelihood) of completed HPV vaccination is much higher in wemen who are white, see an OB-GYN, go to suburban clinics, and do not have medical assistance. Also having a hospital insurance and being in the age group of 11 to 17 are effective parameters that help to complete HPV vaccination. On the other hand, wemen who are black or hispanic, have medical assistance or private Insurance, go to urban clinics, and go to pediatric are less likely to complete HPV vaccination. Therfore, we can conclude that patients of color and low socioeconomic status are less likely to have a completed HPV vaccination. This issue is very obvious when we can see that the white wemon  are twice as likely to complete the HPV vaccine than black wemen. I think all of the variables in the final model makes sense as a strong predictor of our dependent variable and I do not see any unpredictable predictor in our model.

Another intesting result was younger females have a higher completion rate than older females. The only reason that I can guess for this observation is that parents have significant role in completion rate by 
taking care of their children's health. Also the higher rate of completion in suburban clinics than urban clinics is due to the bias that we see in race and socioeconomic status. socioeconomic status  can the main reason of low completion rate in medical assistance insurance. But, the low quality of medical assistance insurance is another point that worth to mention.

One of the problems that I have for making inference out of the data that we have in this lab was the unequal number of wemen in each group (sample size). For example, the number reported Hispanic wemen was mush less than the other gropus of wemen, and changing the size of this group may completly change the result of our investigation.

